---
title: "Project2-Markdown"
output: html_document
date: "2024-01-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Life Expectancy data set PCA analysis.

## Introduction.

### Literature review.

## Data set preliminary review.

### Description of variables.

```{r echo=TRUE}
life = read.csv('Life-Expectancy-Data.csv')
str(life)
head(life)
summary(life)

# dropping the Country, Year, Status columns
life_num = life[, 4:22]
```

```{r}
library(corrplot)
library(psych)
library(zoo)

life_num_filled = na.aggregate(life_num, FUN = mean)
life_num_filled[!complete.cases(life_num_filled),]
```

### Is the data set eligible for factor analysis?

```{r}
life_scaled = scale(life_num_filled)

cor_life = cor(life_scaled)
corrplot(cor_life, method = 'circle', type = "lower", order = "hclust", tl.col = "black", tl.cex = 0.7, number.cex = 0.9)

cortest.bartlett(cor_life)
KMO(cor_life)
```

### Preliminary statistics.

```{r echo=FALSE}
library(kableExtra)
summary(life_num_filled) %>% kbl() %>% kable_paper("hover")
```

```{r echo=FALSE}
library(Hmisc)

life_num_filled = lapply(life_num_filled, as.numeric)
life_num_filled = as.data.frame(life_num_filled)

hist.data.frame(life_num_filled[,1:6])
hist.data.frame(life_num_filled[,7:12])
hist.data.frame(life_num_filled[,12:17])
hist.data.frame(life_num_filled[,18:19])
```

## Principal component analysis.

```{r echo=FALSE}
library(factoextra)

pca = prcomp(life_scaled, center=FALSE, scale=FALSE)
```

### Eigenvalues.

```{r}
fviz_eig(pca, choice = "eigenvalue",  addlabels = TRUE,   main = "Eigenvalues")
```

### Variance.

```{r}
fviz_eig(pca, choice = 'variance', addlabels = TRUE)
```

```{r}
pca_summary = summary(pca)
plot(pca_summary$importance[3,], ylab = 'Cumulative Variance', xlab = 'Component Index', main = 'Cumulative variance carried by components')
```

### Loadings.

```{r}
pca$rotation %>% kbl() %>% kable_paper("hover")
```

```{r}
fviz_pca_var(pca, col.var = "blue")
```

```{r}
fviz_pca_ind(pca, col.ind="contrib", geom = "point", gradient.cols = c("green", "yellow", "red" , "purple", "black"))
```

Contribution visualises how much each variable contributes to creating the principal components. Higher contribution values mean that a variable has a more significant impact on shaping the principal components.

```{r}

fviz_pca_ind(pca, col.ind="cos2", geom = "point", gradient.cols = c("green", "yellow", "red" , "purple", "black"))
```

Cos2 indicates how well individual observations are represented in the reduced-dimensional space created by principal components. The squared cosine for each observation on a principal component reflects the proportion of its variance explained by that component. Higher squared cosine values suggest effective representation by the principal components.

```{r}
PC1 <- fviz_contrib(pca, choice = "var", axes = 1)
PC2 <- fviz_contrib(pca, choice = "var", axes = 2)
PC3 <- fviz_contrib(pca, choice = "var", axes = 3)
PC4 <- fviz_contrib(pca, choice = "var", axes = 4)
PC5 <- fviz_contrib(pca, choice = "var", axes = 5)

fviz_contrib(pca, choice = "var", axes = 1)
fviz_contrib(pca, choice = "var", axes = 2)
fviz_contrib(pca, choice = "var", axes = 3)
fviz_contrib(pca, choice = "var", axes = 4)
fviz_contrib(pca, choice = "var", axes = 5)

# gridExtra::grid.arrange(PC1, PC2, PC3, PC4, PC5, ncol = 3, nrow=2)

```

## References.

<https://www.ibm.com/docs/en/spss-statistics/28.0.0?topic=detection-kmo-bartletts-test>

Shrestha, Anil & Luo, Wei. (2017). Analysis of Groundwater Nitrate Contamination in the Central Valley: Comparison of the Geodetector Method, Principal Component Analysis and Geographically Weighted Regression. ISPRS International Journal of Geo-Information. 6. 297. 10.3390/ijgi6100297.

Thomson RG, De Brún A, Flynn D, et al. Factors that influence variation in clinical decision-making about thrombolysis in the treatment of acute ischaemic stroke: results of a discrete choice experiment. Southampton (UK): NIHR Journals Library; 2017 Jan. (Health Services and Delivery Research, No. 5.4.) Appendix 5, Factor analysis of Institutional Culture Scale. Available from: <https://www.ncbi.nlm.nih.gov/books/NBK410188/>

Abdi, H. and Williams, L.J. (2010) Principal Component Analysis. Wiley Interdisciplinary Reviews: Computational Statistics, 2, 433-459.
